The watershed transform in ITK - discussion and new developments

### 2006-05-23 article/Article.tex
Minor corrections and additions to article.
darcs-hash:20060523061724-fafb9-60e8e3c282a308e5950666fb53147fc8df85152c.gz

Section \ref{sect:usingWT} provides an introduction to a
methodology known as {\em marker based} watershed. Using markers in
watershed transforms is a well established technique in the
morphological image analysis community and has proved very useful for
many applications. The aim of this part of the report is to describe
the technique to a wider audience \footnote{It should also be noted
that the lack of a marker based watershed motivated the recent work on
ITK filter classes.}.

@@

An Obvious alternative that is very similar in practice is to 
smooth the image so that there are fewer regional minima.

@@

This can be achieved by performing a {\em reconstruction by erosion}
or {\em recursive conditional erosion} from the seleted regional
minima. This process is also called {\em minima imposition}. Figure
\ref{fig:recon} illustrates the results of apply a
reconstruction by erosion to a 1D image. The original image is shown
in blue and has regional minima A, B, C, D and E. The marker image
which is recursively eroded is shown in black, with a single zero at
regional minimum B. The red plateaus illustrate the parts of the
result image that are above the orginal. The resulting image has only
on regional minima that corresponds to B.
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.5]{recon}
\caption{Reconstruction by erosion from a single regional minimum.}
\label{fig:recon}
\end{center}
\end{figure}

@@

Early implementations used a special ordered queue for 8 bit data,
which may be where the misconception that a morphological watershed
can only be applied to 8 bit data types arose -- in fact some of the
early work along these lines was covered by a patent\footnote{I'm told
that the associated paper is \cite{Meyer91a} and that there may be an
English version in \cite{Dougherty93a}}. The new ITK watershed
algorithms use standard c++ containers (maps and standard queues) to
implement the ordered queue\footnote{One potential problem with
applying this style of algorithm to non integer data is finding the
regional minima. Reconstruction approaches are often used to find
regional minima of integer type images, but subtracting 1, carrying
out a recursive dilation and subtracting the two. The same concept
does not extend to floating point data. However, if regional minima
are available the algorithms we are discussing can be applied to
floating point data.}.

@@

Algorithm 3 requires slightly more complexity to store label
information in the queue.

@@

\caption{Three difficult test images. The first image, plateau, show that ITK's watershed produce a bad result on this image: only one minima take all the image. This problem is also shown in Figure \ref{fig:cthead1-markers}. The morphological watershed give a better result, but not yet perfect: the separation line should not be curved as it is. This problem is also visible in Figure \ref{fig:embryo-mws}. The next image, thick line, again show that ITK watershed can produce strange results with a green label which is where we should have the light blue one. Morphological watershed gives good results. The last image, button hole, is a very difficult case for watershed algorithms which mark watershed line: the watershed pixel at the bottom of the image stop the label propagation. There is several possible behavior: produce a large watershed line which cover all the zone (Vincent-Soille algorithm, not shown here), let the label on the top of the image propagate in the zone (Meyer algorithm), split the zone in 2 zones with the 2 labels of the bottom of the image (topological watershed, not shown here).\label{fig:difficult-cases}}
\end{center}
\end{figure}

@@

\verb$MorphologicalWatershedFromMarkersImageFilter$ implement the
algorithms \ref{sect:BeucherAlgorithm} and \ref{sect:MeyerAlgorithm}.
They are accessible respectively when the attribute \verb$MarkWatershed$
is \verb$false$ or is \verb$true$. The filter is templated over the
input image type, which can be any scalar type, and the label image type
which should be an integer. The filter take 2 inputs:
\begin{itemize}
  \item a grayscale image which can be set with \verb$SetInput(image)$,
\verb$SetInput(0, image)$ or \verb$SetInput1(image)$.
  \item a marker image which can be set with \verb$SetMarkerImage(image)$
and \verb$SetInput2(image)$. \verb$SetInput(0, image)$ should be avoided
because it will fail if the input image and the marker image are not of the
same type. The markers are all the non-zero values in the image.
\end{itemize}

The algorithm \ref{sect:MeyerAlgorithm} require to keep the status of all
the pixels: {\em already added to the queue} or {\em not yet added to the queue}.
This task is often done by using negative values in the output image
for the pixels not yet added to the queue. In our implementation, we have
chosen to use a binary image to store this, while labels always take values
above 0 and watershed lines are marked with value 0. The negative values
option have several drawbacks:
\begin{itemize}
  \item it requires to use a different image type than the label type defined
by the user in case the user have chosen an unsigned pixel type.
  \item it requires to use one of the value which can be possibly aready used
by a label in the marker image, in case of a signed type.
\end{itemize}
Those problems can be avoided\footnote{It's not true for the unsigned long
pixel type} by using a type which contains a larger range of value than the
type choosed by the user, but then, we must copy the data to an image of
the type chosen by the user at the end of the algorithm.

We prefered using a separated binary image (of type \verb$itk::Image<bool, dim>$)
to store this information. This solution avoid the need to copy the image at
the end, and so should be more efficient in computation time, and require only
one bit per pixel and so should be more memory efficient as well.

@@

\verb$MorphologicalWatershedImageFilter$ perform a watershed transformation
without marker. It encapsulate several filters:
\begin{itemize}
  \item \verb$HMinimaImageFilter$
  \item \verb$RegionalMinimaImageFilter$
  \item \verb$ConnectedComponentImageFilter$
  \item \verb$MorphologicalWatershedFromMarkersImageFilter$
\end{itemize}
While \verb$RegionalMinimaImageFilter$, \verb$ConnectedComponentImageFilter$,
and \verb$MorphologicalWatershedFromMarkersImageFilter$ are the usal steps
of the watershed usage, the \verb$HMinimaImageFilter$ is a convenient way
to avoid the classical watershed's over segmentation by allowing the user
to select the dynamic of the minima. The minimum dynamic of minima can be
selected with \verb$SetLevel()$.

>>>>

Section \ref{sect:usingWT} provides an introduction to a methodology
known as {\em marker based} watershed. Using markers in watershed
transforms is a well established technique in the morphological image
analysis community and has proved very useful for many
applications. The aim of this part of the report is to describe the
technique to a wider audience \footnote{The lack of a marker based
watershed motivated this work on ITK filter classes.}.

@@

An obvious alternative that is very similar in practice is to 
smooth the image so that there are fewer regional minima.

@@

This can be achieved by performing a {\em reconstruction by erosion}
or {\em recursive conditional erosion} from the seleted regional
minima. This process is also called {\em minima imposition}. Figure
\ref{fig:recon} illustrates the result of applying a
reconstruction by erosion to a 1D image. The original image is shown
in blue and has regional minima A, B, C, D and E. The marker image
which is recursively eroded is shown in black, with a single zero at
regional minimum B. The red plateaus illustrate the parts of the
result image that are above the orginal. The resulting image has only
on regional minima that corresponds to B.
\begin{figure}[htbp]
\begin{center}
\includegraphics[scale=0.5]{recon}
\caption{Reconstruction by erosion from a single regional minimum.}
\label{fig:recon}
\end{center}
\end{figure}

@@

Early implementations used a special ordered queue for 8 bit data,
which may be where the misconception that a morphological watershed
can only be applied to 8 bit data types arose -- in fact some of the
early work along these lines was covered by a patent\footnote{I'm told
that the associated paper is \cite{Meyer91a} and that there may be an
English version in \cite{Dougherty93a}}. The new ITK watershed
algorithms use standard c++ containers (maps and standard queues) to
implement the ordered queue\footnote{One potential problem with
applying this style of algorithm to non integer data is finding the
regional minima. Reconstruction approaches are often used to find
regional minima of integer type images, by subtracting 1, carrying out
a recursive dilation and subtracting the two. The same concept does
not extend to floating point data. However, if regional minima are
available the algorithms we are discussing can be applied to floating
point data. It is possible to use other techniques, such as flooding
approaches, to find regional minima of floating point images.}.

@@

@@

\caption{Three difficult test images. The first image, plateau, illustrates a 
bad result produced by ITK's watershed in which most of the area is
assigned to one minima. This problem is also shown in Figure
\ref{fig:cthead1-markers}. The morphological watershed give a better
result, but not yet perfect: the separation line should not be curved
as it is. This problem is also visible in Figure
\ref{fig:embryo-mws}. The reason for the problem is as follows - the 
example is attempting to split a plateau into two regions, and should
therefore produce a Voronoi tesselation, with the two markers being
separated by a straight line. This does not happen because various
approximations have been made, as mentioned in Section \ref{}, that
influence the estimation of the distance function. Even if a more
accurate underlying distance function was used the computations are
still being carried out on a grid and so are limited by the digital
representation of lines.

The next image, thick line, again show that ITK watershed can produce
strange results with a green label which is where we should have the
light blue one. Morphological watershed gives good results. The last
image, button hole, is a very difficult case for watershed algorithms
which mark watershed lines: the watershed pixel at the bottom of the
image stop the label propagation. There are several possible
behaviors: produce a large watershed line which covers all the zone
(Vincent-Soille algorithm, not shown here), let the label on the top
of the image propagate into the zone (Meyer algorithm), split the zone
in 2 zones with the 2 labels of the bottom of the image (topological
watershed, not shown here).\label{fig:difficult-cases}}
\end{center}
\end{figure}

@@

\verb$MorphologicalWatershedFromMarkersImageFilter$ implements
algorithms \ref{sect:BeucherAlgorithm} and \ref{sect:MeyerAlgorithm}.
Setting attribute \verb$MarkWatershedLine$ to \verb$true$ selects
algorithm \ref{sect:MeyerAlgorithm} (Meyer) while setting to
\verb$false$ selects algorithm \ref{sect:BeucherAlgorithm} (Beucher).
The filter is templated over the input image type, which can be any
scalar type, and the label image type which should be an integer. The
filter takes 2 inputs:
\begin{itemize}
  \item a grayscale image which can be set with \verb$SetInput(image)$,
\verb$SetInput(0, image)$ or \verb$SetInput1(image)$.
  \item a marker image which can be set with \verb$SetMarkerImage(image)$
and \verb$SetInput2(image)$. \verb$SetInput(0, image)$ should be avoided
because it will fail if the input image and the marker image are not of the
same type. The markers are all the non-zero values in the image.
\end{itemize}

Algorithm \ref{sect:MeyerAlgorithm} requires the status of all pixels
to be tracked -- pixels are either {\em already added to the queue} or
{\em not yet added to the queue}.  This task is often done by using
negative values in the output image for the pixels not yet added to
the queue. In our implementation, we have chosen to use a binary image
to store this, while labels always take values above 0 and watershed
lines are marked with value 0. The negative values option have several
drawbacks:
\begin{itemize}
  \item it requires to use a different image type than the label type defined
by the user in case the user have chosen an unsigned pixel type.
  \item it requires to use one of the value which can be possibly aready used
by a label in the marker image, in case of a signed type.
\end{itemize}
Those problems can be avoided\footnote{It's not true for the unsigned long
pixel type} by using a type which contains a larger range of value than the
type choosed by the user, but then, we must copy the data to an image of
the type chosen by the user at the end of the algorithm.

{\em Don't think this should be here. Is there any compiler that will
represent bool images as bit vectors? It might save memory but will
increase access time}.  We prefered using a separated binary image (of
type \verb$itk::Image<bool, dim>$) to store this information. This
solution avoid the need to copy the image at the end, and so should be
more efficient in computation time, and require only one bit per pixel
and so should be more memory efficient as well.

@@

\verb$MorphologicalWatershedImageFilter$ performs a watershed transformation
without markers. It encapsulate several filters:
\begin{itemize}
  \item \verb$HMinimaImageFilter$
  \item \verb$RegionalMinimaImageFilter$
  \item \verb$ConnectedComponentImageFilter$
  \item \verb$MorphologicalWatershedFromMarkersImageFilter$
\end{itemize}
While \verb$RegionalMinimaImageFilter$, \verb$ConnectedComponentImageFilter$,
and \verb$MorphologicalWatershedFromMarkersImageFilter$ are the usal steps
of the watershed usage, the \verb$HMinimaImageFilter$ is a convenient way
to avoid the classical watershed's over segmentation by allowing the user
to select the dynamic of the minima. The minimum dynamic of minima can be
selected with \verb$SetLevel()$.
